---
title: "Climbing in Paris"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

# Data processing with R

## Study area preparation : 10 km around Paris

A bounding box of 10 km around Paris (map bounding box). Considering the fact that we are interested here by the accessibility to climbing clubs, a bounding box of 15 km around Paris is also created

IRIS data is imported, Paris is extracted, a buffer of 10km is created to identify the bounding box.

```{r}
library(sf)

# Source : IGN
iris <- st_read("CONTOURS-IRIS.shp", quiet = TRUE)

# Extract Paris
# Département
iris$dep <- substr(iris$INSEE_COM, 1, 2)
paris <- iris[iris$dep == "75",]
paris <- paris[!paris$NOM_IRIS %in% c("Bois de Vincennes 1",
                                      "Bois de Vincennes 2",
                                      "Bois de Boulogne 1",
                                      "Bois de Boulogne 2",
                                      "Bois de Boulogne 3"),]

# Delete Bois-de-Vincennes / Boulogne Iris
paris <- st_union(paris)

# 5 km around Paris map layout
paris5k <- st_buffer(paris, 5000)
paris5k <- st_as_sfc(st_bbox(paris5k, crs = 2154))

# 10 km around Paris (get OSM data) in long/lat
paris10k <- st_buffer(paris, 10000)
paris10k <- st_transform(paris10k, 4326)
paris10k <- st_bbox(paris10k) 
paris10k <- as.vector(paris10k)

# Intersection with IRIS
iris <- st_intersection(iris, paris5k)

# Communes aggregation (layout)
com <- aggregate(iris[,c("INSEE_COM", "NOM_COM")],
                 by = list(iris$INSEE_COM),
                 FUN = head, 1)

# Study area presentation
plot(st_geometry(st_buffer(paris, 10000)), col = "lightgrey", border = NA)
plot(st_geometry(iris), col = "peachpuff", border = "white", lwd = .2, add = TRUE)
plot(st_geometry(com), col = NA, border = "black", lwd = .5, add = TRUE)
```

Data is enriched by socio-economic data (disposible income) for further analysis. We keep only "Habitation" IRIS for origins-destination calculation.

```{r}
library(readxl)
# Import INSEE data
df <- read_xlsx("data-raw/BASE_TD_FILO_DISP_IRIS_2019.xlsx", skip = 5, sheet = "IRIS_DISP")
iris <- merge(iris[,c("CODE_IRIS", "NOM_IRIS", "TYP_IRIS")], df[,c("IRIS", "DISP_MED19")],
              by.x = "CODE_IRIS", by.y = "IRIS", all.x = TRUE)

# Keep only habitation IRIS for origins calculation
ori <- iris[iris$TYP_IRIS == "H",]
ori <- st_centroid(ori)
```

Data is transformed in latitude/longitude for origin-destination calculations and implementation in Observable.

```{r}
# Consolidated data
ori <- st_transform(ori, crs = 4326)
iris <- st_transform(iris, crs = 4326)
com <- st_transform(com, crs = 4326)
```

## Import OSM data with R

Data is imported from OpenStreetMap. We can argue that the accuracy of the data is quite good, since missing points have been added to the OpenStreetMap database with my personal knowledge of this sport activity in Paris and its surroundings.

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
library(osmdata)

# define a bounding box
q0 <- opq(bbox = paris10k) 

# extract climbing areas
q <- add_osm_feature(opq = q0, key = 'sport', value = "climbing")
res <- osmdata_sf(q)
poi <- res$osm_points

# Manage private and associative areas
private <- poi[!is.na(poi$brand),]
asso <- poi[!is.na(poi$federation),]

asso$type <- "Associative structure"
private$type <- "Speculative structure"
poi <- rbind(asso, private)


cols <- c("osm_id", "name", "climbing.toprope", "climbing.boulder", "climbing.length",
          "climbing.routes", "type", "federation", "brand")

poi <- poi[,cols]
```

## Time distance to the nearest climbing area by bike

\[osrm default bicycle profile\](https://github.com/Project-OSRM/osrm-backend/blob/master/profiles/bicycle.lua)

bicycle parameters on \[OSRM\](https://github.com/camptocamp/provelobern_bicyclerouteplanner/wiki/Profiles)

add slope to the calculation : https://www.liedman.net/2015/04/13/add-elevation-data-to-osrm/

mapbox solution : https://blog.mapbox.com/elevation-aware-routing-profiles-in-mapbox-directions-21e182a85165

```{r}
library(osrm)
options(osrm.server = "http://188.166.36.154:5000/", osrm.profile = "bike")

# Manage ids
row.names(ori) <- ori$CODE_IRIS

# Whatever the structure
bike_all <- osrmTable(src = ori, dst = poi, 
                      measure = "duration")
xx <- bike_all$duration

# Name of the nearest structure
osm_id <- colnames(xx)[apply(xx, 1, which.min)] # Name
osm_id <- data.frame(osm_id, stringsAsFactors = FALSE)
osm_id$iris <- row.names(xx)
osm_id <- merge(osm_id, poi[,c("osm_id", "name", "type")], 
                by = "osm_id", all.x = TRUE)
osm_id <- st_set_geometry(osm_id, NULL)

# Time to the nearest climbing area
time <- apply(xx, 1, min) # Time
time <- data.frame(time, stringsAsFactors = FALSE)
time$iris <- row.names(time)
osm_id <- merge(osm_id, time, by = "iris", all.x = TRUE)

# Number of climbing area at less than 10 minutes by bike
n10mn <- xx
n10mn <- data.frame(xx, stringsAsFactors = FALSE)
n10mn[n10mn < 10] <- 1
n10mn[n10mn > 10] <- 0
n10mn$N <- rowSums(n10mn)
n10mn$iris <- row.names(n10mn)
osm_id <- merge(osm_id, n10mn[,c("iris", "N")], by = "iris",
                all.x = TRUE)

osm_id <- osm_id[,c(1,3:4, 6:7)]
colnames(osm_id) <- c("CODE_IRIS", "ALL_NAME", "TYPE", "ALL_TIME", "ALL_10MN")
iris <- merge(iris, osm_id, by = "CODE_IRIS", all.x = TRUE)



# Prive climbing club (fees $$$)
poi_priv <- poi[poi$type == "Speculative structure",] 

bike_priv <- osrmTable(src = ori, dst = poi_priv, 
                      measure = "duration")
xx <- bike_priv$duration

# Name of the nearest structure
osm_id <- colnames(xx)[apply(xx, 1, which.min)] # Name
osm_id <- data.frame(osm_id, stringsAsFactors = FALSE)
osm_id$iris <- row.names(xx)
osm_id <- merge(osm_id, poi[,c("osm_id", "name", "type")], 
                by = "osm_id", all.x = TRUE)

# Time to the nearest climbing area
time <- apply(xx, 1, min) # Time
time <- data.frame(time, stringsAsFactors = FALSE)
time$iris <- row.names(time)
osm_id <- merge(osm_id, time, by = "iris", all.x = TRUE)

# Number of climbing area at less than 10 minutes by bike
n10mn <- xx
n10mn <- data.frame(xx, stringsAsFactors = FALSE)
n10mn[n10mn < 10] <- 1
n10mn[n10mn > 10] <- 0
n10mn$N <- rowSums(n10mn)
n10mn$iris <- row.names(n10mn)
osm_id <- merge(osm_id, n10mn[,c("iris", "N")], by = "iris",
                all.x = TRUE)
osm_id <- osm_id[,c(1,3, 6:7)]
colnames(osm_id) <- c("CODE_IRIS", "PRIV_NAME", "PRIV_TIME", "PRIV_10MN")
iris <- merge(iris, osm_id, by = "CODE_IRIS", all.x = TRUE)
  

# Associative structure - FFME
poi_asso <-poi[poi$type == "Associative structure",] 
poi_ffme <- poi_asso[poi_asso$federation == "FFME",]

bike_ffme <- osrmTable(src = ori, dst = poi_ffme, 
                      measure = "duration")
xx <- bike_ffme$duration

# Name of the nearest structure
osm_id <- colnames(xx)[apply(xx, 1, which.min)] # Name
osm_id <- data.frame(osm_id, stringsAsFactors = FALSE)
osm_id$iris <- row.names(xx)
osm_id <- merge(osm_id, poi[,c("osm_id", "name", "type")], 
                by = "osm_id", all.x = TRUE)

# Time to the nearest climbing area
time <- apply(xx, 1, min) # Time
time <- data.frame(time, stringsAsFactors = FALSE)
time$iris <- row.names(time)
osm_id <- merge(osm_id, time, by = "iris", all.x = TRUE)

# Number of climbing area at less than 10 minutes by bike
n10mn <- xx
n10mn <- data.frame(xx, stringsAsFactors = FALSE)
n10mn[n10mn < 10] <- 1
n10mn[n10mn > 10] <- 0
n10mn$N <- rowSums(n10mn)
n10mn$iris <- row.names(n10mn)
osm_id <- merge(osm_id, n10mn[,c("iris", "N")], by = "iris",
                all.x = TRUE)
osm_id <- osm_id[,c(1,3, 6:7)]
colnames(osm_id) <- c("CODE_IRIS", "FFME_NAME", "FFME_TIME", "FFME_10MN")
iris <- merge(iris, osm_id, by = "CODE_IRIS", all.x = TRUE)

# Associative structure - FSGT
poi_fsgt <- poi_asso[poi_asso$federation == "FSGT",]

bike_fsgt <- osrmTable(src = ori, dst = poi_fsgt, 
                      measure = "duration")
xx <- bike_fsgt$duration

# Name of the nearest structure
osm_id <- colnames(xx)[apply(xx, 1, which.min)] # Name
osm_id <- data.frame(osm_id, stringsAsFactors = FALSE)
osm_id$iris <- row.names(xx)
osm_id <- merge(osm_id, poi[,c("osm_id", "name", "type")], 
                by = "osm_id", all.x = TRUE)

# Time to the nearest climbing area
time <- apply(xx, 1, min) # Time
time <- data.frame(time, stringsAsFactors = FALSE)
time$iris <- row.names(time)
osm_id <- merge(osm_id, time, by = "iris", all.x = TRUE)

# Number of climbing area at less than 10 minutes by bike
n10mn <- xx
n10mn <- data.frame(xx, stringsAsFactors = FALSE)
n10mn[n10mn < 10] <- 1
n10mn[n10mn > 10] <- 0
n10mn$N <- rowSums(n10mn)
n10mn$iris <- row.names(n10mn)
osm_id <- merge(osm_id, n10mn[,c("iris", "N")], by = "iris",
                all.x = TRUE)
osm_id <- osm_id[,c(1,3, 6:7)]
colnames(osm_id) <- c("CODE_IRIS", "FSGT_NAME", "FSGT_TIME", "FSGT_10MN")
iris <- merge(iris, osm_id, by = "CODE_IRIS", all.x = TRUE)


# A function to display results in this iris dataframe
st_write(iris, "data-conso/iris.geojson")
st_write(com, "data-conso/com.geojson")
st_write(poi, "data-conso/poi.geojson")
```
